Revealing transient strain in geodetic data with Gaussian process regression

### 2017-11-08 RevisedSubmission/hines_hetland_2017_gji.tex
did more

% issues and potential improvements on noise

Our results demonstrate that GPR is an effective tool for estimating
transient strain rates and detecting geophysical processes from GNSS
data. However, there are some aspects of our method that warrant
further research. For example, there is potential for a more thorough
analysis of the spatio-temporal noise in the GNSS data than what was
performed in Section \ref{sec:NoiseModel}. Specifically, we did not
explore spatially correlated noise, such as from common mode error. We
have assumed that any spatially correlated noise has a sufficiently
long wavelength that it has a negligible effect on transient strain
rates. We can also improve upon the seasonal noise model used in this
study, which consists of four spatially uncorrelated sinusoids for
each station. We did not explore the spatial covariance of seasonal
deformation or the temporal roughness (i.e., the number of sinusoids
needed to describe the observations). The periodic Gaussian process
\citep{Mackay1998} is an alternative model for seasonal deformation
and is well suited for exploring the roughness of seasonal
deformation.  The periodic Gaussian process has zero mean and the
covariance function

\begin{equation}\label{eq:Periodic}
  T(t,t') = \phi^2 \exp\left(\frac{-\sin(\pi|t - t'|)^2}{2\tau^2}\right). 
\end{equation}

Realizations have annual periodicity and the roughness is controlled
by $\tau$. Decreasing $\tau$ has the same effect as including higher
frequency sinusoids in the seasonal model. The optimal value for
$\tau$ can be found with the REML method as described in Section
\ref{sec:NoiseModel}.

@@

There are numerical considerations as well ...
The transient strain rates estimated in this study are constrained by
about seven years of daily displacement observations from 94 GNSS
stations. It can be computationally intensive to evaluate eqs.
(\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) for a dataset
with this size. We significantly reduce the amount of memory needed to
estimate transient strain rates by describing the temporal covariance
of displacements with a compact Wendland covariance function. Using a
compact covariance function for our prior turns eqs.
(\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) into sparse
systems of equations, which we then solve with CHOLMOD. CHOLMOD is
designed for solving sparse, positive definite systems of equations.
The matrix being inverted in eqs. (\ref{eq:PosteriorMean3}) and
(\ref{eq:PosteriorCov3}) is not positive definite; however, we can use
another partitioned matrix inversion identity from \citet{Press2007}
to partition it into positive definite submatrices to be inverted.
Even when using a compact covariance function, it may still be
necessary to reduce the computational burden by dividing the data into
subsets and evaluating transient strain rates for each subset.

>>>>

%% Areas for improvement and further study
% noise model

Our results demonstrate that GPR is an effective tool for estimating
transient strain rates and detecting geophysical processes from GNSS
data. However, there are some aspects of our method that warrant
further research. For example, there is potential for a more thorough
analysis of the spatio-temporal noise in the GNSS data than what was
performed in Section \ref{sec:NoiseModel}. Specifically, we did not
explore spatially correlated noise, such as common mode error. We
have assumed that any spatially correlated noise has a sufficiently
long wavelength that it has a negligible effect on transient strain
rates. We can also improve upon the seasonal noise model used in this
study, which consists of four sinusoids for each station. We did not
explore the roughness of the seasonal deformation (i.e., the number of
sinusoids needed to describe the deformation). The periodic Gaussian
process \citep{Mackay1998} is an alternative model for seasonal
deformation and is well suited for exploring the roughness of seasonal
deformation.  The periodic Gaussian process has zero mean and the
covariance function
\begin{equation}\label{eq:Periodic}
  T(t,t') = \phi^2 \exp\left(\frac{-\sin(\pi|t - t'|)^2}{2\tau^2}\right). 
\end{equation}
Realizations have annual periodicity and the roughness is controlled
by $\tau$. Decreasing $\tau$ has the same effect as including higher
frequency sinusoids in the seasonal model. The optimal value for
$\tau$ can be found with the REML method.

@@

% numerical issues

Another potential research direction would be to reduce the
computational cost of our method for estimating transient strain
rates. GPR is generally computationally expensive when there are many
observations. The transient strain rates estimated in this study are
constrained by seven years of daily displacement observations from 94
GNSS stations, which amounts to about 240,000 observations for each
displacement component. For a dataset with this size, it is difficult
to evaluate the matrix inverses in eqs. (\ref{eq:PosteriorMean3}) and
(\ref{eq:PosteriorCov3}). We alleviate this computational cost by
using a compact Wendland covariance function for our prior. By using a
compact covariance function for our prior, eqs.
(\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) become sparse
systems of equations that are more tractable to evaluate. However, the
sparsity decreases when the Wendland covariance function has a larger
time-scale parameter. The Wendland covariance then offers less of a
computational advantage when we were interested in geophysical
processes that occur over longer times-scales. An alternative approach
to dealing with large datasets would be to explore approximation
methods for GPR \citep[sec. 8]{Rasmussen2006}.
